---
title: "SQL databases and R"
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float: yes
    font:family: Lato
  pdf_document:
    toc: yes
---

Learning Objectives:

- Access a database from *R*.
- Run SQL queries using `RSQLite` and `dplyr`.
- Describe the lazy behaviour of `dplyr` on data stored in a database outside of *R*.
- Prototype queries and retrieve all final results.
- Create complex queries across on of multiple databse tables.
- Create an SQLite database from existing `.csv` files.

# The `portal_mammals` database

We can explore the `surveys` data. First, we load in the `dbplry` and `RSQLite' package.

```{r}
library(dplyr)
library(dbplyr)
library(RSQLite)
```

Let's now download our file `portal_mammals.sqlite`. 

```{r}
dir.create("raw_data", showWarnings = FALSE)
download.file(url = "https://ndownloader.figshare.com/files/2292171",
              destfile = "raw_data/portal_mammals.sqlite", mode = "wb")
```

# Connecting to databases

We can point `R` to this database.

```{r}
mammals <- DBI::dbConnect(RSQLite::SQLite(), "raw_data/portal_mammals.sqlite")
```

This command uses two packages enable `dbplry` and `dplyr` communicate with the SQLite database. "DBI" is not something that we'll ever use directly as a user. It allows *R* to send commands to databses irrespetive of the database management system used. The `RSQLite` package allows *R* to interface with SQLite databases.

This command does not load the data into the *R* session (unlike the `read_csv()` function). Instead, it merely instructs *R* tp connect to the SQLite database contained in the `portal_mammals.sqlite` file.

Let's now take a closer look at the `mammals` database we've connected toc

```{r}
src_dbi(mammals)
```

Just like a spreadsheet with multiple worksheets, a SQLite database can contain multiple tables. In this case, three of them are listed in the `tbls` ros in the output above:

- plots
- species
- surveys
  
Now that we know we can connect to the databse, let's explore how to get the data from its tables in *R*.

## Querying the database with the `SQL` syntax

To connect to tables within a database, we can use the `tbl()` function from `dplyr`. This function can be used to send SQL queries to the database. To demonstrate this, let's select the columns `year`, `species_id`, and `plot_id` from the surveys table.

```{r}
tbl(mammals, sql("SELECT year, species_id, plot_id FROM surveys"))
```

## Querying the database with the `dplyr` syntax

One of the strengths of `dplyr` is that the same operations can be done using `dplyr`'s verbs instead of writing `SQL`. First, we select the table on which to do the operations by creating the surveys object, and then we use the same standard `dplyr` syntax as if it were a data frame.

```{r}
surveys <- tbl(mammals, "surveys")
surveys %>%
  select(year, species_id, plot_id)
```

In this case, the `surveys` object behaves like a data frame. Several functions that can be used with data frames can also be used on tables from a database. For instance, the `head()` function can be used to check the first 10 rows of the table.

```{r}
head(surveys, 10)
```

This output looks just like a regular dataframe. The table has 9 columns, and the `head()` function shows us the furst 10 rows. Note that the columns `plot_type`, `taxa`, `genus`, and `species` are missing. These are now located in the tables `plots` and `species` which we will join together in a moment.

However, some functions don't work quite as expected. For instance, let's check how many rows are in total using `nrow()`

```{r}
nrow(surveys)
```

That's strange - *R* doesn't know how many rows the `surveys` table contains, it returns `NA` instead. If we have a look at the first output from the `head()` function, we notice that the first line includes '???`, indicating that the number of rows wasn't known. 

The reason for this behaviour highlights a key difference between using `dplyr` on datasets in memory (e.g. loaded into your *R* session via `read_csv()`), and also provided by a database. To understand it, we take a closer look at how `dplyr` communicates with our `SQLite` database.

## SQL Translation

Relational databases typically use a special-purpose language, `Strcutured Query Language (SQL), to manage and query data.

For example, the following SQL query returns the first 10 rows from the surveys table

` 
SELECT 
FROM 'surveys' 
LIMIT 10
`

Behind the scenes, `dplyr`:

- Translates the *R* code in *SQL*.
- Submits it to the database.
- Tranlates the database's response into an *R* data frame.

To see exactly what's going on, we can use `dplyr`'s `show_query()` function to show which `SQL` commands are actually sent to the database:

```{r}
show_query(head(surveys, 10))
```

The output shows the actual SQL query sent to the database, it matches our manually constructed `SELECT` statement above. 

Instad of having to formulate the SQL query ourselves, and having to mentally switch back and forth between *R* and *SQL* syntax, we can delegate this translation to `dplyr`. The best bit is, we don't even have to know SQL to interact with a database via `dplyr`.

`dplyr`, in turn, doesn't do the real work of subsetting the table, either. Instead, it sends the query to the database, waits for its response and returns it to us.

That way, *R* never gets to see the full `surveys` table - and that's why it could not tell us how many rows it contains. However, this allows us to work with large datasets - even too large to fit into our computer's memory.

`dplyr` can translate many different query types into SQL, allowing us to e.g. `select()` specific columns, `filter()` rows, or join tables.

To see this in action, let's compose a few queries with `dplyr`

## Simple Database Queries

First, let's request rows of the surveys table in which `weight` < 5, and keep only the `species_id`, `sex`, and `weight` columns. 

```{r}
surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)
```

Executing this command will return a table with 10 rows and the required `species_id`, `sex`, and `weight`. Perfect.

However, there appears to be only 10 rows. Why did *R* only retrieve 10 rows?

# Laziness

When working with `dplyr` it tried to the minimal work required:

- It never pulls data into *R* unless you specifically ask it. 
- It delays doing any work until the last possible moment - it collects together everything you want to do and then it sends it to the database in one step. 

When you construct a `dplyr` query, you can connect multiple verbs into a single pipeline. For example, together everything you want to do and then sends it to the database in one step. 

If we wanted to, we could add on even more steps - eg remove the `sex` column as an additional `select` call:

```{r}
data_subset <- surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)

data_subset %>%
  select(-sex)
```

Just like the first `select()` call, the `select(-sex)` command is not executed by *R*. It is sent to the database instead. Only the *final result* is retrieved.

We could always add on more steps, e.g. we could `filter()` by `species_id` or minimum `weight`. That's why *R* doesn't retrieve the full set of results - instead, it only retrieves the first 10 results from the database by default (After all, you might want to add an additional step and get the database to do more work...)

To instruct,, *R* to stop being lazy e.g. to retrieve all of the query results from the database, we add the `collect()` function to our pipe. It indicates that our database query is finished, time to get the *final* results and load them into the *R* session.

```{r}
data_subset <- surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight) %>%
  collect()
```

Now we have all 17 rows that match our query in a dataframe, and can continue to work with the exclusviely in *R*, without communicating with the database. 

# Complex database qeuries

`dplyr` enables database queries across one or multiple database tables, using the same single and multiple table verbs we encountered previously. This means we can use the same commands regardless of whether you interact with a remote database or local dataset. This is a really useful feature if you work with large datasets: we can first prototype our code on a small subset that fits into memory, and when your code is ready; you can change the input datasets to your full database without having to change the syntax. 

On the other hand, being able to use `SQL` queries can be useful if your collaborators have already put together complex queries to prepare the data that we need for our analysis.

To illustrate how to use `dplyr` with these complex queries, we are going to join the `plots and `surveys` tables. The `plots` table in the database contains information about the different plots surveyed by the researchers. To access it, we point the `tbl()` command to it.

```{r}
plots <- tbl(mammals, "plots")
plots
```

The `plot_id` column also features in the `surveys` table

Because `plot_id` is listed in both tables, and we can use it to look up matching records, and join the two tables.

If we have two tables named x and y with a common `ID`, we can join them using the `join` functions, two of which are described and illustrated below.

- `inner_join()`: This returns all rows from x where they are matching values in y, and all columns from x and y.
- `left_join()`: This returns all rows from x, and all columns from x and y. Rows in x with no match in y will have `NA` values in the new_columns.

In both forms of join, if there are multiple matches between x and y, all combinations of the matches are returned. For the full list of `join()` functions, check out the `tidyverse` join page.

For example, to extract all surveys from the first plot, which has `plot_id`, we can do:

```{r}
plots %>%
  filter(plot_id == 1) %>%
  inner_join(surveys) %>%
  collect()
```

Without the `collecet()` statment, only the first 10 matching rows are returned. By adding `collect()`, the full set of 1,985 rows are retrieved. 

# Challenge

Let's write a query that returns the number of rodents observed in each plot in each year.

```{r}
head(surveys, 10)
surveys %>%
  filter(species_id == "")
```