---
title: "SQL databases and R"
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float: yes
    font:family: Lato
  pdf_document:
    toc: yes
---

Learning Objectives:

- Access a database from *R*.
- Run SQL queries using `RSQLite` and `dplyr`.
- Describe the lazy behaviour of `dplyr` on data stored in a database outside of *R*.
- Prototype queries and retrieve all final results.
- Create complex queries across on of multiple databse tables.
- Create an SQLite database from existing `.csv` files.

# The `portal_mammals` database

We can explore the `surveys` data. First, we load in the `dbplry` and `RSQLite' package.

```{r}
library(dplyr)
library(dbplyr)
library(RSQLite)
```

Let's now download our file `portal_mammals.sqlite`. 

```{r}
dir.create("raw_data", showWarnings = FALSE)
download.file(url = "https://ndownloader.figshare.com/files/2292171",
              destfile = "raw_data/portal_mammals.sqlite", mode = "wb")
```

# Connecting to databases

We can point `R` to this database.

```{r}
mammals <- DBI::dbConnect(RSQLite::SQLite(), "raw_data/portal_mammals.sqlite")
```

This command uses two packages enable `dbplry` and `dplyr` communicate with the SQLite database. "DBI" is not something that we'll ever use directly as a user. It allows *R* to send commands to databses irrespetive of the database management system used. The `RSQLite` package allows *R* to interface with SQLite databases.

This command does not load the data into the *R* session (unlike the `read_csv()` function). Instead, it merely instructs *R* tp connect to the SQLite database contained in the `portal_mammals.sqlite` file.

Let's now take a closer look at the `mammals` database we've connected toc

```{r}
src_dbi(mammals)
```

Just like a spreadsheet with multiple worksheets, a SQLite database can contain multiple tables. In this case, three of them are listed in the `tbls` ros in the output above:

- plots
- species
- surveys
  
Now that we know we can connect to the databse, let's explore how to get the data from its tables in *R*.

## Querying the database with the `SQL` syntax

To connect to tables within a database, we can use the `tbl()` function from `dplyr`. This function can be used to send SQL queries to the database. To demonstrate this, let's select the columns `year`, `species_id`, and `plot_id` from the surveys table.

```{r}
tbl(mammals, sql("SELECT year, species_id, plot_id FROM surveys"))
```



